prolfqua&colon; A Comprehensive R-package for Proteomics Differential Expression Analysis

### 2022-10-26 inst/MyArticle/paper.Rmd
added changes from MdE.

The peptide abundances (from the MaxQuant _peptide.txt_ file) were $log_2$ transformed and subsequently scaled, where median and the mean average deviation (mad) were obtained from the human proteins only. We removed _one hit wonders_, i.e., proteins with a single peptide assignment. Protein abundances are inferred from the peptide intensities using Tukey's median polish. Finally, we fitted the fixed effect linear models, the dropout model `r BiocStyle::Biocpkg("proDA")` to protein abundances, the mixed effect linear model, the `r BiocStyle::Biocpkg("PECA")` model, and hurdle model `r BiocStyle::Biocpkg("msqrob2")` to peptide intensities.

@@

@valikangas2016systematic discuss and benchmark various methods of peptide or protein intensity normalization, such as variance stabilizing normalization [@huber2002variance] or quantile normalization [@bolstad2003comparison]. In this work, we use a robust version of the z-score, where instead of the mean $\bar{x}$ we use the median, and instead of the standard deviation $\tilde{S}$ we use the median absolute deviation (mad):

@@

Given a linear model $y = \beta X$, we can compute the difference $\beta_{c}$ between two groups by a linear combination of weights $c$ and model parameters $\beta$, where $c$ is a column vector with as many elements as there are coefficients $\beta$ in the linear model. If $c$ has $0$ for one or more of its rows, then the corresponding coefficient in $\beta$ is not involved in determining the contrast [@ph525xseries].

@@

Given the weights $c$ contrasts from model parameter estimates obtained with `r BiocStyle::Biocpkg('proDA')`, or `r BiocStyle::Biocpkg('msqrob2')` can be determined, by this enabling us to call these packages using the same application programming interface (API).

@@

However, if there is no observation in a group, the model fit fails. For example, suppose a protein is unobserved in all the samples of a group. In that case, a plausible explanation is that the protein abundance is below the limit of detection (LOD) of the MS instrument. In such a case, we will substitute the group mean, using the expected protein abundance $A$ at the LOD $A_{LOD}$.
To estimate $A_{LOD}$ we are using the protein abundances of those groups where the protein was observed in only a single sample. Typically there are many such cases, and hence we take the median.

>>>>

The peptide abundances (from the MaxQuant _peptide.txt_ file) were $log_2$ transformed and subsequently scaled, where median and the mean absolute deviation were obtained from the human proteins only. We removed _one hit wonders_, i.e., proteins with a single peptide assignment. Protein abundances are inferred from the peptide intensities using Tukey's median polish. Finally, we fitted the fixed effect linear models, the dropout model `r BiocStyle::Biocpkg("proDA")` to protein abundances, the mixed effect linear model, the `r BiocStyle::Biocpkg("PECA")` model, and hurdle model `r BiocStyle::Biocpkg("msqrob2")` to peptide intensities.

@@

@valikangas2016systematic discuss and benchmark various methods of peptide or protein intensity normalization, such as variance stabilizing normalization [@huber2002variance] or quantile normalization [@bolstad2003comparison]. In this work, we use a robust version of the z-score, where instead of the mean we use the median $\tilde{x}$ , and instead of the standard deviation we use the median absolute deviation $\tilde{S}$:

@@

Given a linear model $y = \beta X$, we can compute the difference $\beta_{c}$ between two groups by the dot product of weights $c$ and model parameters $\beta$, where $c$ is a column vector with as many elements as there are coefficients $\beta$ in the linear model. If $c$ has $0$ for one or more of its rows, then the corresponding coefficient in $\beta$ is not involved in determining the contrast [@ph525xseries].

@@

Given the weights $c$ contrasts from model parameters obtained with `r BiocStyle::Biocpkg('proDA')`, or `r BiocStyle::Biocpkg('msqrob2')` can be determined, which will enable us to implement adapters to `proDA` or `msqrob2`.

@@

However, if there is no observation in a group, the model fit fails. For example, suppose a protein is unobserved in all the samples of a group. In that case, a plausible explanation is that the protein abundance is below the limit of detection (LOD) of the MS instrument. In such a case, we will substitute the group mean, using the expected protein abundance $A$ at the LOD $A_{LOD}$.
To estimate $A_{LOD}$ we are using the protein abundances of those groups where the protein was observed in only a single sample (see Supplement section Estimating $A_{LOD}$). Typically there are many such cases, and hence we take the median.
